CFTP Convergence

### 2025-01-23 application_examples/abc_inferences/archipelago-definition.Rmd
Updating changes in archipelago-definition

if (!requireNamespace("GenomicRanges", quietly = TRUE)) {
  BiocManager::install("GenomicRanges")
}

library(GenomicRanges)
library(rtracklayer)
```

@@

## Step 1: Definition of promoter regions

@@

For that, we use [Annotation Features File (GTF) from the NCBI RefSeq Assembly](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/)

```{r}
# Load the GTF file
gtf_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gtf" 
gtf <- import(gtf_file)
head(gtf)
```

The gtf file is structured as a GRanges object containing genomic annotation information with the following key components (columns):

- `seqnames`: Chromosome or scaffold name (e.g., NC_036780.1).

- `ranges`: Start and end positions of the genomic feature.

- `strand`: Strand of the feature (+ for forward strand, - for reverse strand).

- `source`: Annotation source (e.g., Gnomon).

- `type`: Feature type (e.g., gene, transcript, exon).

- `Metadata`: Additional columns include gene_id, transcript_id, gene, product, gene_biotype, exon_number, etc., which describe the feature in detail.

Example Entries:

- Row 1: A gene on the + strand, identified as LOC106676409, classified as lncRNA.

- Row 2: A transcript corresponding to the same gene.

- Rows 3-5: Exons belonging to the transcript.

- Row 6: A gene on the - strand, classified as protein_coding.

We filter for transcripts:

```{r}
# Filter for transcripts
gtf_transcripts <- gtf[gtf$type == "transcript", ]
head(gtf_transcripts)
```

And define the TSS by accounting for the strand orientation so that:

- For Features on the + Strand: The TSS is the smallest genomic coordinate (start).

- For Features on the - Strand: The TSS is the largest genomic coordinate (end), as transcription occurs in the reverse direction.

```{r}
# Extract TSS
promoters <- data.frame(
  chr = seqnames(gtf_transcripts),
  tss = ifelse(strand(gtf_transcripts) == "+", start(gtf_transcripts), end(gtf_transcripts)),
  strand = strand(gtf_transcripts)
)
head(promoters)
```

The promoters data frame contains:

- `chr`: Chromosome names

- `tss`: Transcription start sites

- `strand`: Strand orientation of the transcripts

@@

Convert the promoter and CGI data to `GRanges` objects for genomic interval operations:

```{r}
# Create GRanges for promoters
promoter_gr <- GRanges(
  seqnames = promoters$chr,
  ranges = IRanges(start = promoters$tss, end = promoters$tss),
  strand = promoters$strand
)

# Create GRanges for CGI
cgi_gr <- GRanges(
  seqnames = cgif$chr,
  ranges = IRanges(start = cgif$start, end = cgif$end)
)
```

Use `findOverlaps` to find promoters within or near CpG islands:

```{r}
# Find overlaps
overlaps <- findOverlaps(promoter_gr, cgi_gr)

# Extract overlapping promoter indices
promoters_with_cgi <- promoters[queryHits(overlaps), ]
cgi_overlaps <- cgif[subjectHits(overlaps), ]

# Combine data for output
promoter_cgi_overlap <- cbind(promoters_with_cgi, cgi_overlaps)
head(promoter_cgi_overlap)
```

The `promoter_cgi_overlap` dataset combines information about promoters and CpG islands (CGIs) where they overlap. It contains the following columns:

- `chr`: The chromosome where the promoter is located.

- `tss`: The transcription start site (TSS) position of the promoter.

- `strand`: The strand of the promoter (+ or -).

- `chr`: The chromosome where the CGI is located (same as the promoter in the overlap context).

- `start`: The start position of the overlapping CGI.

- `end`: The end position of the overlapping CGI.

- `length`: The length of the CGI in base pairs.

- `CpGcount`: The number of CpG dinucleotides within the CGI.

- `GCcount`: The number of G and C nucleotides within the CGI.

- `pctGC`: The percentage of G and C nucleotides within the CGI.

- `obsExp`: The observed-to-expected ratio of CpG dinucleotides in the CGI.

Each row corresponds to an overlap between a promoter and a CGI. For example:

- The first row shows that a promoter at TSS 755,445 on the negative strand overlaps with a CGI on chromosome NC_036780.1 spanning positions 755,397 to 756,689. This CGI has a length of 1,293 bp, contains 123 CpG dinucleotides, and has a GC content of 63.19% with an observed-to-expected CpG ratio of 0.954.

- The overlap is detailed for both the promoter and the CGI, including their respective properties.

```{r}
#write.table(promoter_cgi_overlap, file = "promoter_cgi_overlap.txt", sep = "\t", row.names = FALSE)
promoter_cgi_overlap <- read.table("/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/promoter_cgi_overlap.txt")
```

### Check TSS overlap density

```{r}
plot(
  promoter_cgi_overlap$length, 
  promoter_cgi_overlap$pctGC, 
  xlab = "CGI Length (bp)", 
  ylab = "GC Content (%)", 
  main = "Relationship Between CGI Length and GC Content",
  pch = 16,  # Solid circle points
  col = "blue"  # Point color
)

@@

```{r}

>>>>

library(rtracklayer)
```

@@

We will perform this steps for the genomic region of interest:

```{r}
dpp4_chr <- "NC_036801.1"
dpp4_start <- 29502362
dpp4_end <- 29517382
region_start <- dpp4_start - 5e5
region_end <- dpp4_end + 5e5
```

- DPP4 gene location:

  - Chromosome: NC_036801.1

      - Start: 29,502,362

      - End: 29,517,382

- Extended Region: Includes 500,000 base pairs upstream and downstream of the DPP4 gene.

## Step 1: Definition of "promoter/upstream" regions

@@

### Choice of gene annotiation file format

File formats from [NCBI](https://www.ncbi.nlm.nih.gov/datasets/genome/GCF_000238955.4/). Selected RefSeq only (no Ensembl annotation). More info [here](https://www.biobam.com/differences-between-gtf-and-gff-files-in-genomic-data-analysis/), [here](http://www.ensembl.org/info/website/upload/gff.html) and [here](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-020-6707-9).

- The General Feature Format (GFF): GFF is a standard file format used for storing genomic sequences and annotations, developed by the Sanger Centre (v2) and the Sequence Ontology Project (v3). It describes features and annotations of DNA, RNA, and protein sequences in a plain text format with nine fields separated by tabs. Focuses on the hierarchical structure of genome annotations and is often less verbose.

- The Gene Transfer Format (GTF): GTF, often seen as a derivative of GFF, is more specific in its application. It was generally termed GTF around 2000 and developed as part of collaborative genome annotation projects. GTF borrows from GFF but has additional structures for detailed gene annotations, often used to describe transcripts and their coding sequences.

The `.gtf` file contains more information, but for our purpose we can use the `.gff` file.

```{r}
# Load the GFF file
gff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gff" 
gff <- as.data.frame(import(gff_file))

- `seqnames`: Chromosome.

- `start`, `end`: Coordinates of the gene.

- `strand`: Gene orientation.

- `type`: Feature type (here filtered for genes).

- `Gene`: Gene name or ID.

- `gene_biotype`: Functionality (e.g., protein_coding)

```{r}
table(gff$source)
```

- `source`: indicates the source of the annotation for each feature in the file. These sources represent the tools, pipelines, or databases that generated the corresponding annotation.

  - `RefSeq`: These annotations come directly from the RefSeq database. RefSeq annotations are high-quality, manually curated or computationally derived features that are part of NCBI's Reference Sequence project. Examples: Protein-coding genes, RNA genes, pseudogenes.

   - `Gnomon`: Gnomon is NCBI's gene prediction tool, which generates computationally predicted annotations. These annotations may include protein-coding genes, non-coding RNAs, or other features predicted based on sequence analysis. Gnomon is used when there is no manually curated annotation available for certain regions or genes.

   - `cmsearch`: This refers to the Infernal package (specifically the cmsearch tool), which identifies RNA features such as non-coding RNAs or structured RNA elements.  Often used for annotating RNA families based on covariance models.

  - `tRNAscan-SE`: tRNAscan-SE is a specialized tool for identifying and annotating tRNA genes in genome sequences. Annotations with this source indicate regions identified as tRNAs through this tool.

   - `BestRefSeq`: Features with this source are part of the BestRefSeq annotation, which integrates high-quality manual curation with computational predictions to produce a "best" annotation for a gene or transcript. BestRefSeq annotations are typically the most reliable.

  - BestRefSeq,Gnomon (or `BestRefSeq%2CGnomon`): This indicates that a feature is annotated by both the BestRefSeq and Gnomon pipelines. Such annotations may combine curated data with computational predictions.

### Filter region of interest

```{r}
region_gff <- gff[
  gff$seqnames == "NC_036801.1" & 
  gff$type == "gene" & 
  gff$start <= region_end & 
  gff$end >= region_start, 
]

head(region_gff)
```

```{r}
table(region_gff$source)
```

```{r}
region_gff[grep("GeneID:101483164", region_gff$Dbxref), ]
```

```{r}
length(unique(region_gff$ID))
unique(region_gff$ID)
```

```{r}
# Extract relevant columns
gene_data <- region_gff[, c("strand","start", "end", "gene")]

# Create a y-axis index for unique gene names
gene_data$y_index <- match(gene_data$gene, unique(gene_data$gene))

# Set gene colors based on strand information (red for negative strand, blue for positive strand)
gene_data$gene_color <- ifelse(gene_data$strand == "-", "red", "blue")

```{r}
# Set up larger plotting area and adjust margins
par(mar=c(6, 10, 4, 2), xpd=TRUE)

# Create the plot with adjusted y-axis
plot(c(min(gene_data$start), max(gene_data$end)), 
     c(0, max(gene_data$y_index) + 1), 
     type="n", 
     xlab="Genomic Coordinates", 
     ylab="", 
     xaxt="n", 
     yaxt="n")

# Plot each gene as a line, with thicker line for GeneID:101483164
for (i in 1:nrow(gene_data)) {
  # Set line width for GeneID:101483164 (2x as thick)
  lwd_value <- ifelse(gene_data$gene[i] == "LOC101483164", 4, 1)

  # Plot the gene line with the appropriate color and line width
  segments(gene_data$start[i], gene_data$y_index[i], gene_data$end[i], gene_data$y_index[i], 
           col=gene_data$gene_color[i], lwd=lwd_value)
}

# Add gene names to the y-axis
axis(2, at=gene_data$y_index, labels=gene_data$gene, las=1, cex.axis=0.7)
# Add x-axis with genomic coordinates
axis(1)

# Optionally, add a grid for better visualization
grid()

# Add a legend at the top-left corner
legend("topleft", 
       legend=c("Positive strand (blue)", "Negative strand (red)"), 
       fill=c("blue", "red"), 
       bty="n", cex=0.8)
```

```{r}
# Sort gene_data by strand and then by start position
gene_data_sorted <- gene_data[order(gene_data$strand, gene_data$start), ]

# Initialize a vector to store overlap results
gene_data_sorted$overlap <- FALSE

# Loop over genes, comparing each gene with the next one in the same strand
for (i in 1:(nrow(gene_data_sorted) - 1)) {
  # Check if genes are on the same strand and overlap
  if (gene_data_sorted$strand[i] == gene_data_sorted$strand[i + 1] &&
      gene_data_sorted$end[i] >= gene_data_sorted$start[i + 1]) {
    gene_data_sorted$overlap[i] <- TRUE
    gene_data_sorted$overlap[i + 1] <- TRUE
  }
}

# View the genes with overlap information
head(gene_data_sorted)

```{r}
gene_data_sorted[which(gene_data_sorted$overlap==TRUE),"gene"]
```

```{r}
region_gff[region_gff$gene == "nop58",]
region_gff[region_gff$gene == "LOC111501074",]
```

The second one corresponds to the source `cmsearch`, so it may be the one that should be taken out.

```{r}
region_gff <- region_gff[region_gff$gene != "LOC111501074",]
```

### Define the 200 kb upstream regions

Calculate the upstream start and end to prevent overlap with the previous gene. *This is done for each strand from the second gene onwards as we dont know where the gene before the first gene is as it is not included in the region dataframe*.

- For genes on the positive strand, the upstream region starts 20kb before the gene's start position, or at the end of the previous gene if there is one.

- For genes on the negative strand, the upstream region starts 20kb after the gene's end position, or at the start of the next gene if there is one.

```{r}
# Separate genes by strand
positive_strand_genes <- region_gff[region_gff$strand == "+", ]
negative_strand_genes <- region_gff[region_gff$strand == "-", ]

# Sort genes by start position within each strand
positive_strand_genes <- positive_strand_genes[order(positive_strand_genes$start), ]
negative_strand_genes <- negative_strand_genes[order(negative_strand_genes$end, decreasing = TRUE), ]

# Initialize the new columns for upstream start and end
positive_strand_genes$upstream_start <- NA
positive_strand_genes$upstream_end <- NA

negative_strand_genes$upstream_start <- NA
negative_strand_genes$upstream_end <- NA

# Calculate the upstream region for positive strand genes
for (i in 2:nrow(positive_strand_genes)) {
  positive_strand_genes$upstream_start[i] <- max(positive_strand_genes$start[i] - 20000, positive_strand_genes$end[i-1] + 1)
  positive_strand_genes$upstream_end[i] <- positive_strand_genes$start[i] - 1
}

# Calculate the upstream region for negative strand genes
for (i in 2:nrow(negative_strand_genes)) {
  negative_strand_genes$upstream_start[i] <- min(negative_strand_genes$end[i] + 20000, negative_strand_genes$start[i-1] - 1)
  negative_strand_genes$upstream_end[i] <- negative_strand_genes$end[i] + 1
}

# Combine the results back
region_gff <- rbind(positive_strand_genes, negative_strand_genes)

# View the updated dataframe with upstream regions
head(region_gff)

#### Chromosome lengths

In case the upstream regions were to be defined for all chromosomes/scaffolds, the chromosome lengths would be needed to be the maximum for the start of the upstream region of the genes on the negative strand.

Load chromosome lengths:

```{r, eval = FALSE}
gbff_file <- "/home/sara/methylation/MethEvolSIM/application_examples/abc_inferences/genomic_dist/ncbi_dataset/ncbi_dataset/data/GCF_000238955.4/genomic.gbff" 
file_lines <- readLines(gbff_file)
# Extract the LOCUS line
locus_lines <- file_lines[grep("^LOCUS", file_lines)]
head(locus_lines)
```

They don't seem to end with the dot and the number (version) as in the case of the `gff` data. Lets check if the `gff` has several versions:

```{r}
length(unique(gff$seqnames))
# Remove version suffix from seqnames
length(unique(sub("\\.\\d+$", "", gff$seqnames)))
```

It does not have several versions.

```{r, eval = FALSE}
# Extract chromosome names and lengths
chrom_info <- gsub("^LOCUS\\s+(\\S+)\\s+(\\d+)\\s+.*$", "\\1,\\2", locus_lines)
chrom_info <- strsplit(chrom_info, ",")
chrom_names <- sapply(chrom_info, `[`, 1)
chrom_lengths <- as.numeric(sapply(chrom_info, `[`, 2))
# Create a named vector
chrom_lengths <- setNames(chrom_lengths, chrom_names)

# Remove version numbers from seqnames in the region-filtered gff
region_gff$seqnames <- sub("\\.\\d+$", "", region_gff$seqnames)

@@

```{r}
region_cgi <- cgif[cgif$chr==dpp4_chr,]
```

```{r}
# Create a matrix of CGI intervals for faster lookup
cgi_intervals <- cbind(region_cgi$start, region_cgi$end)

# Function to efficiently check if positions are within any CGI
is_within_cgi_vector <- function(positions, cgi_intervals) {
  # Apply the check for each position against all intervals
  sapply(positions, function(pos) {
    any(pos >= cgi_intervals[,1] & pos <= cgi_intervals[,2])
  })
}

# Initialize results list for better memory management
results_list <- vector("list", nrow(region_gff))

# Process each gene
for (i in 1:nrow(region_gff)) {
  upstream_start <- region_gff$upstream_start[i]
  upstream_end <- region_gff$upstream_end[i]

  # Skip if NA values
  if (is.na(upstream_start) | is.na(upstream_end)) {
    next
  }

  # Create sequence of positions (handling both forward and reverse cases)
  if (upstream_start > upstream_end) {
    positions <- seq(from = upstream_start, to = upstream_end, by = -1)
  } else {
    positions <- seq(from = upstream_start, to = upstream_end, by = 1)
  }

  # Get CGI status for all positions at once
  in_cgi <- is_within_cgi_vector(positions, cgi_intervals)

  # Calculate distances properly based on strand
  distances <- if(region_gff$strand[i] == "+") {
    # For positive strand: distance from gene start
    region_gff$start[i] - positions  # Positive values will be upstream
  } else {
    # For negative strand: distance from gene end
    positions - region_gff$end[i]    # Positive values will be upstream
  }

  # Create result data frame for this gene
  results_list[[i]] <- data.frame(
    gene_id = region_gff$gene[i],
    position = positions,
    in_cgi = in_cgi,
    distance = distances,
    strand = region_gff$strand[i]
  )
}

# Combine all results efficiently
upstream_data <- do.call(rbind, results_list[!sapply(results_list, is.null)])

# Calculate counts and probabilities
position_counts <- table(upstream_data$distance)  # Using table() for correct counts
distance_seq <- as.numeric(names(position_counts))

# Calculate probabilities
cgi_probabilities <- tapply(upstream_data$in_cgi, 
                          upstream_data$distance, 
                          mean)   # Probability of being in CGI

@@

```{r, fig.height=12, fig.width=10}
# Set up the plot with better formatting
par(mfrow = c(2,1))  # Set up 2 plots vertically
par(mar = c(4, 5, 4, 2))  # Adjust margins for top plot
# Plot 1: Probability plot
plot(distance_seq, cgi_probabilities, 
     xlab = "",  # Remove x-label from top plot
     ylab = "Probability of Position Being in CGI",
     main = "Probability of Position Being in CpG Island\nby Distance to Gene Start",
     cex.lab = 1.2,      # Larger axis labels
     cex.axis = 1.1,     # Larger axis numbers
     cex.main = 1.3,     # Larger title
     las = 1,            # Horizontal axis labels
     pch = 16,          # Solid circle points
     ylim = c(0, 1),    # Force probability y-axis to [0,1]
     col = adjustcolor("darkblue", alpha.f = 0.5))  # Semi-transparent blue

# Add grid
grid(col = "gray90", lty = "dotted")

# Add a horizontal line at y = 0
abline(h = 0, col = "gray70", lty = 2)

# Plot 2: Count plot
par(mar = c(5, 5, 3, 2))  # Adjust margins for bottom plot
plot(distance_seq, position_counts, type = "h",  # Using the correct counts
     xlab = "Distance to Gene Start (bp)",
     ylab = "Number of Positions",
     main = "Number of Positions at Each Distance",
     col = "darkgreen",
     lwd = 1,
     ylim = c(0, max(position_counts) * 1.05))  # Force y-axis to start at 0 with 5% padding
grid(col = "gray90", lty = "dotted")
```

```{r, fig.height=4, fig.width=15}
# Set up plotting area with adjusted margins
par(mfrow = c(1,1))  
par(mar=c(5, 8, 4, 4))  # Increased right margin for legend

# Create y-coordinates for each type
y_coords <- c(1, 2, 3)  # 1 for neg strand, 2 for pos strand, 3 for CGIs

# Get the x-axis range including both genes and CGIs
x_range <- range(c(region_gff$start, region_gff$end))

# Create the empty plot
plot(x_range, c(0.5, 3.5), 
     type="n", 
     xlab="Genomic Coordinates",
     ylab="",
     yaxt="n",  # Remove y-axis ticks
     main="Genes and CpG Islands")

# Add custom y-axis labels
axis(2, at=y_coords, 
     labels=c("- Strand Genes", "+ Strand Genes", "Islands"),
     las=1)  # Horizontal labels

# Plot genes on negative strand
neg_genes <- region_gff[region_gff$strand == "-", ]
for(i in 1:nrow(neg_genes)) {
    segments(neg_genes$start[i], y_coords[1], 
            neg_genes$end[i], y_coords[1],
            col="red", lwd=2)
}

# Plot genes on positive strand
pos_genes <- region_gff[region_gff$strand == "+", ]
for(i in 1:nrow(pos_genes)) {
    segments(pos_genes$start[i], y_coords[2], 
            pos_genes$end[i], y_coords[2],
            col="blue", lwd=2)
}

# Plot CGIs
region_cgis <- cgif[cgif$chr == dpp4_chr & 
                   cgif$start >= region_start & 
                   cgif$end <= region_end,]

for(i in 1:nrow(region_cgis)) {
    segments(region_cgis$start[i], y_coords[3], 
            region_cgis$end[i], y_coords[3],
            col="darkgreen", lwd=2)
}

# Add grid
grid(col="gray90")

### 2025-01-23 extensive_tests/CFTP_convergence/draft-test_CFTP_convergence.Rmd
Updating last changes to be able to pull
Updating changes in archipelago-definition

# CFTP

@@

## Implementation

Currently neglects IWEs but allows to sample a sequence of methylation states from the equilibrium given by the equilibrium frequencies of the genomic regions, the correlated and independent SSE.

## Test

This workflow uses [Snakemake](https://snakemake.readthedocs.io) to manage the steps to test that the CFTP method samples a sequence of methylation states from the equilibrium given by the model accounting for the correlated SSE process (while neglecting the IWE process).

For that it samples a number of parameter combinations and for each it simulates initial data for a number of replicates. It uses as summary statistics the mean correlation of a central segment of methylation states per genomic structure type. We expect that the value of this summary statistics increases with the increase of evolutionary time until it converges towards the equilibrium given by the corresponding parameter combination. We track its value by discretizing the evolution along the branch in small steps (length as config.yaml file value for `branch_length`). Then, we call the CFTP method on the initial data to sample the sequence of methylation states from the equilibrium and compute its corresponding value for the summary statistics. We expect the distribution of the summary statistics obtained by calling the CFTP method to be representative of the equilibrium obtained after the branch evolution has reached the equilibrium.

**Add summary and graphical image**

### Requirements

@@

When using nohup one needs to activate the environment in the command.

@@

## Steps in the Workflow

@@

## Results

```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("your caption"), echo=FALSE}
knitr::include_graphics("Figures/CFTP_testConvergence_paramsID_01.pdf")
```

>>>>

```{r setup}
#library(MethEvolSIM)
library(devtools)
load_all()
```

**Update way in which MethEvolSIM is loaded**

# Introduction

@@

**Add examples of use**

CFTP method is defaulted to FALSE but can be set to TRUE when simulating initial data or data evolution along a tree.

Sample initial data with the sequence of methylation states according to the equilibrium.

```{r}
# Example with sampled initial methylation frequencies and default parameter values
custom_infoStr <- data.frame(n = c(100, 100, 100),
                             globalState = c("M", "U", "M"))
initialD <- simulate_initialData(infoStr = custom_infoStr, CFTP = TRUE)
```

Simulate evolution of data along a tree calling the CFTP method to sample a sequence of methylation states at the root of the tree from the equilibrium.

```{r}
# Example with customized initial methylation frequencies, customized parameter values 
# and default dt (0.01)
tree <- "((a:1, b:1):2, c:2, (d:3.7, (e:4, f:1):3):5);"
custom_infoStr <- data.frame(n = c(10, 10, 10),
                             globalState = c("M", "U", "M"),
                             u_eqFreq = c(0.1, 0.8, 0.1),
                             p_eqFreq = c(0.1, 0.1, 0.1),
                             m_eqFreq = c(0.8, 0.1, 0.8))
custom_params <- get_parameterValues()
custom_params$mu <- 0.005

evolD <- simulate_evolData(infoStr = custom_infoStr, tree = tree, params = custom_params, CFTP = TRUE)
```

## Visual Examples

To ease understanding of how applying CFTP results in sampling an initial sequence of methylation states from the equilibrium towards which the data would evolve when initial data neglects local correlations, we provide the plots obtained while testing the method. The test was performed on 10 parameter combinations as the model equilibrium is dependent on those. It uses as summary statistics the mean correlation of a central segment of methylation states per genomic structure type (see vignette for summary statistics **add link or reference or exact name here**). When simulating without using CFTP, the initial sequence of methylation states is given by the independent sampling of the methylation state of the sites from the equilibrium frequencies of the genomic structure to which they belong. Then, the local correlations between the sites methylation states build up with evolutionary time.

Therefore, we expect that the value of the increases with the increase of evolutionary time until it converges towards the equilibrium given by the corresponding parameter combination. We track its value by discretizing the evolution along a tree branch (representing linear time) in small steps (of length 0.01). Additionally, we call the CFTP method on the initial data to sample the sequence of methylation states from the equilibrium and compute its corresponding value for the summary statistics. We expect the distribution of the summary statistics obtained by calling the CFTP method to be representative of the equilibrium obtained after the branch evolution has reached the equilibrium.

In the following figures, the y axis represents the value of the mean correlation summary statistics and the x axis represents the number of discretized time steps. Each figure corresponds to one of the 10 sampled parameter combinations and for each parameter combination 10 replicates were simulated, represented with the different colors. Dots (with a different shape for each of the replicates) represent the value of the mean correlation at the given time step and the dashed line represents its value after calling CFTP on the initial data.

![Sampled parameter combination 1](Figures/CFTP_testConvergence_paramsID_01.png)

![Sampled parameter combination 2](Figures/CFTP_testConvergence_paramsID_02.png)

![Sampled parameter combination 3](Figures/CFTP_testConvergence_paramsID_03.png)

![Sampled parameter combination 4](Figures/CFTP_testConvergence_paramsID_04.png)

![Sampled parameter combination 5](Figures/CFTP_testConvergence_paramsID_05.png)

![Sampled parameter combination 6](Figures/CFTP_testConvergence_paramsID_06.png)

![Sampled parameter combination 7](Figures/CFTP_testConvergence_paramsID_07.png)

![Sampled parameter combination 8](Figures/CFTP_testConvergence_paramsID_08.png)

![Sampled parameter combination 9](Figures/CFTP_testConvergence_paramsID_09.png)

![Sampled parameter combination 10](Figures/CFTP_testConvergence_paramsID_10.png)

### Reproducibility

The test used to generate the visual example is provided as a workflow, which can be run using [Snakemake](https://snakemake.readthedocs.io) or by running the scripts separatedly. Here, we describe the structure and content of the Snakemake workflow. However, the scripts that comprise the workflow are available under `extensive_tests/CFTP_convergence/scripts`.

#### Workflow Requirements

@@

When using nohup one can activate the environment as in the following command.

@@

#### Steps in the Workflow

@@
